---
title: "Fig2"
author: "James Drew"
date: "19/02/2021"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## The Moffitt et al. 2021 dataset 

Expression data is taken from the series matrix file. Data has been preprocessed as per: 	log2 Cy5 signal was analyzed after setting negative values to zero. Probe data were collapsed by gene symbol via mean expression. Samples were quantile normalized to the subset of genes with less than 0.5 standard deviation across primary, metastatic, and normal samples.

```{r}
# load dependencies
options(connectionObserver = NULL) # this is added to fix a bug with loading  ‘org.Hs.eg.db’ pkg in Rstudio
library(org.Hs.eg.db)
library(GSEABase)
library(GSVA)
library(ggplot2)
library(tidyr)
library(limma)
library(gplots)
library(dplyr)
library(RColorBrewer)
library(DESeq2)
library(ggpubr)
```

## Data cleanup
```{r}
# read in data
setwd <- getwd()
c1_Hallmark_Sets <- getGmt('h.all.v7.2.entrez.gmt')
load('HALLMARK_INTERFERON_genelist.RData')
load('KEGG_MYC_TARGETS_genelist.RData')
moffit.expr <- read.delim2('GSE71729_series_matrix.txt', skip=26)

# keep primary tumor, liver, lung and LN met samples
sample_filter <- grepl(pattern = 'Primary.Pancreas|Met.Liver|Met.LymphNode|Met.Lung',colnames(moffit.expr))
sample_filter[1] <- TRUE
moffit.expr <- moffit.expr[,sample_filter]

# split metadata and expr
moffit.meta <- moffit.expr[1:42,]
moffit.expr <- moffit.expr[43:nrow(moffit.expr),]
colnames(moffit.expr) <- moffit.expr[1,]
moffit.expr <- moffit.expr[-c(1),]
rownames(moffit.expr) <- moffit.expr$ID_REF
moffit.expr <- moffit.expr[,-c(1)]

# convert to numeric
moffit.expr <- apply(moffit.expr,c(1,2),as.numeric)

# replace na with 0
print(paste(sum(is.na(moffit.expr))," genes removed because NA"))
moffit.expr[is.na(moffit.expr)] <- 0

# update symbol names
entrez.list <- alias2SymbolTable(rownames(moffit.expr), species="Hs") # find updated symbol names
print(paste(sum(is.na(entrez.list))," genes with no symbol map and ",sum(duplicated(entrez.list))," duplicated symbols"))
# clean matrix
moffit.expr <- moffit.expr[!is.na(entrez.list)&!duplicated(entrez.list),]
entrez.list <- entrez.list[!is.na(entrez.list)&!duplicated(entrez.list)]
gene.list <- AnnotationDbi::select(org.Hs.eg.db, keys = entrez.list, column = c("SYMBOL","ENTREZID"), keytype = "SYMBOL")
gene.list <- gene.list[!duplicated(gene.list$SYMBOL),]

# update row names
rownames(moffit.expr) <- gene.list$SYMBOL
```
FIG2B: HEATMAP OF MYC-INF GENES
```{r}
## order samples by tissue site
liver.samples <- grepl(pattern = "Liver",colnames(moffit.meta)[2:ncol(moffit.meta)])
lung.samples <- grepl(pattern = "Lung",colnames(moffit.meta)[2:ncol(moffit.meta)])
lymph.samples <- grepl(pattern = "Lymph",colnames(moffit.meta)[2:ncol(moffit.meta)])
pancreas.samples <- grepl(pattern = "Pancreas",colnames(moffit.meta)[2:ncol(moffit.meta)])
moffit.expr.ordered <- moffit.expr[,c(which(liver.samples),which(lung.samples),which(lymph.samples),which(pancreas.samples))]
# compute full geneset arrays
moffitt.myc2.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% myc_targets_v2,]
moffitt.ifna.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_alpha,]
moffitt.ifng.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_gamma,]
moffitt.gsea.expr <- rbind(moffitt.myc2.expr,moffitt.ifna.expr, moffitt.ifng.expr)
# PLOT FIG2B
pdf(file="fig2b.pdf")
heatmap.2(moffitt.gsea.expr,
        Rowv=NA,
        Colv = T,
        margins = c(5,20),
        key=FALSE,
        cexRow = 1,
        cexCol = 0.8,
        col=colorRampPalette(c("purple", "white", "green"))(299),
        breaks= c(seq(-5,-0.3,length=100),seq(-0.29,0.29,length=100),seq(0.3,5,length=100)),
        scale="row",
        dendrogram = "col",
        density.info="none",
        trace="none",
        RowSideColors = c(rep('green',nrow(moffitt.myc2.expr)),rep('red',nrow(moffitt.ifna.expr)),rep('blue',nrow(moffitt.ifng.expr))),
        ColSideColors = c(rep('green',sum(liver.samples)), rep('purple',sum(lung.samples)),rep("yellow",sum(lymph.samples)),rep("grey",sum(pancreas.samples))))
dev.off()

## order samples by moffit classification
no.samples <-grepl(pattern = ": 0",moffit.meta[19,2:ncol(moffit.meta)])
classical.samples <- grepl(pattern = ": 1",moffit.meta[19,2:ncol(moffit.meta)])
basal.samples <- grepl(pattern = ": 2",moffit.meta[19,2:ncol(moffit.meta)])
moffit.expr.ordered <- moffit.expr[,c(which(no.samples),which(classical.samples),which(basal.samples))]
# compute full geneset arrays
moffitt.myc2.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% myc_targets_v2,]
moffitt.ifna.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_alpha,]
moffitt.ifng.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_gamma,]
moffitt.gsea.expr <- rbind(moffitt.myc2.expr,moffitt.ifna.expr, moffitt.ifng.expr)
# plot
classificationcols <- c(rep('grey',sum(no.samples)), rep('black',sum(classical.samples)),rep("white",sum(basal.samples)))
heatmap.2(moffitt.gsea.expr,
        Rowv=NA,
        Colv = T,
        margins = c(5,20),
        key=T,
        cexRow = 1,
        cexCol = 0.8,
        col=colorRampPalette(c("purple", "white", "green"))(299),
        breaks= c(seq(-5,-0.3,length=100),seq(-0.29,0.29,length=100),seq(0.3,5,length=100)),
        scale="row",
        dendrogram = "none",
        density.info="none",
        trace="none",
        RowSideColors = c(rep('green',nrow(moffitt.myc2.expr)),rep('red',nrow(moffitt.ifna.expr)),rep('blue',nrow(moffitt.ifng.expr))),
        ColSideColors = classificationcols,
        symm=F,symkey=F,symbreaks=T)


## order samples by stromal activity
stroma.na <-grepl(pattern = ": 0",moffit.meta[20,2:ncol(moffit.meta)])
stroma.low <-grepl(pattern = ": 1",moffit.meta[20,2:ncol(moffit.meta)])
stroma.normal <-grepl(pattern = ": 2",moffit.meta[20,2:ncol(moffit.meta)])
stroma.activated <-grepl(pattern = ": 3",moffit.meta[20,2:ncol(moffit.meta)])
moffit.expr.ordered <- moffit.expr[,c(which(stroma.na),which(stroma.low),which(stroma.normal),which(stroma.activated))]
# compute full geneset arrays
moffitt.myc2.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% myc_targets_v2,]
moffitt.ifna.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_alpha,]
moffitt.ifng.expr <- moffit.expr.ordered[rownames(moffit.expr.ordered) %in% hallmark_ifn_gamma,]
moffitt.gsea.expr <- rbind(moffitt.myc2.expr,moffitt.ifna.expr, moffitt.ifng.expr)
# plot heatmap
classificationcols <- c(rep('black',sum(stroma.na)), rep('grey',sum(stroma.low)),rep("white",sum(stroma.normal)), rep("red",sum(stroma.activated)))
heatmap.2(moffitt.gsea.expr,
        Rowv=NA,
        Colv = T,
        margins = c(5,20),
        key=T,
        cexRow = 1,
        cexCol = 0.8,
        col=colorRampPalette(c("purple", "white", "green"))(299),
        breaks= c(seq(-5,-0.3,length=100),seq(-0.29,0.29,length=100),seq(0.3,5,length=100)),
        scale="row",
        dendrogram = "none",
        density.info="none",
        trace="none",
        RowSideColors = c(rep('green',nrow(moffitt.myc2.expr)),rep('red',nrow(moffitt.ifna.expr)),rep('blue',nrow(moffitt.ifng.expr))),
        ColSideColors = classificationcols,
        symm=F,symkey=F,symbreaks=T)
```
## FIG2C: MYC-INF signature in MET SITES
```{r}
# LIVER
liver.filt <- grepl(pattern = "Liver|Pancreas",colnames(moffit.meta)[2:ncol(moffit.meta)])
moffit.expr.liver <- moffit.expr[,liver.filt]
moffit.meta.liver <- moffit.meta[,c(TRUE,liver.filt)]
# run GSVAs on different genesets
data.input <- moffit.expr.liver
rownames(data.input) <- gene.list$ENTREZID # change to entrez ids for GSVA
moffit.gsva.liver <- gsva(expr=data.input, 
                         gset.idx.list=c1_Hallmark_Sets, 
                         method="gsva", 
                         kcdf="Gaussian", 
                         min.sz=10, 
                         max.sz=250)
# setup linear model
TissueSite <- factor(as.character(moffit.meta.liver[15,2:ncol(moffit.meta.liver)]))
gr <- TissueSite == "Liver_Metastasis"
des <- model.matrix(~gr)
data.fit.liver <- lmFit(moffit.gsva.liver, design = des)
data.fit.liver <- eBayes(data.fit.liver)
data.top.liver <- topTable(data.fit.liver, number = Inf, sort.by= "p")


# LYMPH
lymph.filt <- grepl(pattern = "Lymph|Pancreas",colnames(moffit.meta)[2:ncol(moffit.meta)])
moffit.expr.lymph <- moffit.expr[,lymph.filt]
moffit.meta.lymph <- moffit.meta[,c(TRUE,lymph.filt)]
# run GSVAs on different genesets
data.input <- moffit.expr.lymph
rownames(data.input) <- gene.list$ENTREZID # change to entrez ids for GSVA
moffit.gsva.lymph <- gsva(expr=data.input, 
                         gset.idx.list=c1_Hallmark_Sets, 
                         method="gsva", 
                         kcdf="Gaussian", 
                         min.sz=10, 
                         max.sz=250)
# setup linear model
TissueSite <- factor(as.character(moffit.meta.lymph[15,2:ncol(moffit.meta.lymph)]))
gr <- TissueSite == "LymphNode_Metastasis"
des <- model.matrix(~gr)
data.fit.LN <- lmFit(moffit.gsva.lymph, design = des)
data.fit.LN <- eBayes(data.fit.LN)
data.top.LN <- topTable(data.fit.LN, number = Inf, sort.by= "p")

# LUNG
lung.filt <- grepl(pattern="Lung|Primary",colnames(moffit.meta)[2:ncol(moffit.meta)])
moffit.expr.lung <- moffit.expr[,lung.filt]
moffit.meta.lung <- moffit.meta[,c(TRUE,lung.filt)]
# run GSVAs on different genesets
data.input <- moffit.expr.lung
rownames(data.input) <- gene.list$ENTREZID # change to entrez ids for GSVA
moffit.gsva.lung <- gsva(expr=data.input, 
                         gset.idx.list=c1_Hallmark_Sets, 
                         method="gsva", 
                         kcdf="Gaussian", 
                         min.sz=10, 
                         max.sz=250)
# setup linear model
TissueSite <- factor(as.character(moffit.meta.lung[15,2:ncol(moffit.meta.lung)]))
gr <- TissueSite == "Lung_Metastasis"
des <- model.matrix(~gr)
data.fit.lung <- lmFit(moffit.gsva.lung, design = des)
data.fit.lung <- eBayes(data.fit.lung)
data.top.lung <- topTable(data.fit.lung, number = Inf, sort.by= "p")


## FIG2C PLOT
data.top.liver.sig <- rownames(data.top.liver)[data.top.liver$adj.P.Val<0.05 & abs(data.top.liver$logFC) > 0.2]
data.top.lung.sig <- rownames(data.top.lung)[data.top.lung$adj.P.Val<0.05 & abs(data.top.lung$logFC) > 0.2]
data.top.LN.sig <- rownames(data.top.LN)[data.top.LN$adj.P.Val<0.05 & abs(data.top.LN$logFC) > 0.2]
hallmark.sig.list <- unique(c(data.top.liver.sig, data.top.lung.sig, data.top.LN.sig))
hallmark.sig.logfc <- data.frame(geneset = hallmark.sig.list,
                                 liver = data.top.liver[hallmark.sig.list,'logFC'],
                                 lung = data.top.lung[hallmark.sig.list,'logFC'],
                                 lymphnode = data.top.LN[hallmark.sig.list,'logFC'])
hallmark.sig.logfc <- hallmark.sig.logfc %>% arrange(desc(liver))
hallmark.sig.logfc$geneset <- factor(hallmark.sig.logfc$geneset,levels=hallmark.sig.logfc$geneset)

pdf(file="fig2c.pdf")
hallmark.sig.logfc %>% 
  pivot_longer(cols = c('liver','lung','lymphnode'), values_to = "logFC") %>%
  ggplot(aes(x=name,y=geneset)) +
  geom_point(aes(size=abs(logFC), colour=(logFC>0))) + 
  scale_y_discrete(name ="", labels=hallmark.sig.logfc$geneset)+
  theme_classic()
dev.off()
```

